Notes on Validated Model Counting \\ Version of \today

### 2022-03-28 doc/mc-notes.tex
Fixed typo
New notation

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
%\newcommand{\tautology}{1}
%\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

\newcommand{\interp}{\alpha}
\newcommand{\interpset}[1]{{\cal M}(#1)}
\newcommand{\mcount}{\mu}

@@

@@

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemmas~\ref{lemma:negation}--\ref{lemma:sum}, along with the definition of the ITE operator.

>>>>

\newcommand{\one}{\mbox{\bf 1}}
\newcommand{\zero}{\mbox{\bf 0}}
\newcommand{\leafone}{L_1}
\newcommand{\leafzero}{L_0}
\newcommand{\booland}{\land}
\newcommand{\boolor}{\lor}
\newcommand{\boolxor}{\oplus}
\newcommand{\boolnot}{\neg}
%\newcommand{\tautology}{\top}
%\newcommand{\nil}{\bottom}
\newcommand{\tautology}{1}
\newcommand{\nil}{0}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\ite}{\mbox{\it ITE}}

@@

\newcommand{\interp}{\alpha}
\newcommand{\interpset}[1]{{\cal M}(#1)}
\newcommand{\ring}{{\cal Z}}
\newcommand{\cost}{\sigma}
\newcommand{\density}{\rho}
\newcommand{\hashset}{{\cal H}}
\newcommand{\fhash}{h}
\newcommand{\mcount}{\mu}

@@

Let $X = \{x_1, x_2, \ldots, x_n\}$ be a set of Boolean variables.  An
{\em assignment} is a function $\interp$ assigning Boolean values to
the variables: $\interp:X \rightarrow \{\nil, \tautology\}$.  We can
also view an assignment as a set of {\em literals} $\{l_1, l_2,
\ldots, l_n\}$, where each literal $l_i$ is either $x_i$ or
$\obar{x}_i$, corresponding to the assignments $\interp(x_i)$ = 1 or 0,
respectively.

\subsection{Boolean Functions}

A {\em Boolean function} $f:2^X \rightarrow \{0,1\}$ can be
characterized by the set of assignments for which the function
evaluates to 1: $\interpset{f} = \{ \interp | f(\interp) = 1\}$.  Let
$\one$ denote the Boolean function that assigns value 1 to every
assignment, and $\zero$ denote the assignment that assigns value 0 to
every assignment.  These are characterized by the universal and empty
assignment sets, respectively.

From
this we can define the {\em negation} of function $f$ as the function
$\boolnot f$ such that
$\interpset{\boolnot f} = \{ \interp | f(\interp) = 0\}$.
We can also define the conjunction and disjunction operations over functions $f_1$ and $f_2$ as characterized by the sets
$\interpset{f_1 \booland f_2} = \interpset{f_1} \cap \interpset{f_2}$ and
$\interpset{f_1 \boolor f_2} = \interpset{f_1} \cup \interpset{f_2}$.

For assignment $\interp$ and a Boolean formula $E$ over $X$, we
use the notation $\interp\subs{x_i}{E}$ to denote the assignment
$\interp'$, such that $\interp'(x_j) = \interp(x_j)$ for all $j \not = i$
and $\interp'(x_i) = \interp(E)$, where $\interp(E)$ is the value obtained by evaluating formula $E$ with each variable assigned the value given by $\interp$.
In particular, the notation
$\interp\subsflip{x_i}$ indicates the assignment in which the value
assigned to $x_i$ is complemented, while others remain unchanged.

A Boolean function $f$ is said to be {\em independent} of variable
$x_i$ if every $\interp \in \interpset{f}$ has $\interp\subsflip{x_i}
\in \interpset{f}$.  The {\em dependency set} of $f$, denoted
$\depend(f)$ consists of all variables $x_i$ for which $f$ is {\em
  not} independent.

\subsection{Separable Cost Functions}

Let $\ring$ denote the elements of a commutative ring.  A {\em
  separable cost function} $\cost:X \rightarrow \ring$ assigns a value
from the ring to each variable.  We extend this function by defining
the cost of literal as $\cost(\obar{x}_i) = 1 - \cost(x_i)$, the cost
of an assignment as $\cost(\interp) = \prod_{l_i \in \interp}
\cost(l_i)$, and the cost of a function $f$ as $\cost(f) =
\sum_{\interp \in \interpset{f}} \cost(\interp)$.

{\bf Example 1}: Let $\ring$ be the set of rational numbers and
$\cost(x_i) = 1/2$ for all variables $x_i$.  Then cost of every
assignment will be $1/2^{n}$, and the cost of a function will be its
{\em density}, denoted $\density(f)$.  That is $\density(f) \in [0,
  1]$ will be the fraction of assignments for which the function
evaluates to 1.  Given an ability to compute the density of
function $f$, we can scale this by $2^n$ to compute $|\interpset{f}|$, the core task of model counting.
Using density as our metric has the advantage that it
does not vary when the function is embedded in a larger domain $X'
\supseteq X$.

{\bf Example 2}: Let $\ring$ be a field with $|\ring| > 2n$,
and let $\hashset$ be the set functions
mapping elements of $X$ to elements of $\ring$.  For
two distinct functions $f_1$ and $f_2$ and a randomly chosen $\fhash
\in \hashset$, the probability that $h(f_1) = h(f_2)$ will be at most
$2^n/|\ring|< 1/2$.  Therefore, these functions can be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.

\subsection{Computing Cost Functions}

Three key properties of separable cost functions makes it
possible, in some cases, to compute the cost of a Boolean formula without enumerating all of its satisfying solutions.

\begin{lemma}[Complementation]
  For separable cost function $\cost$ and Boolean function $f$:
  $\cost(\neg f) = 1 - \cost(f)$.
\end{lemma}

\begin{lemma}[Conjunction over Independent Domains]
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\depend(f_1) \cap \depend(f_2) = \emptyset$:
    $\cost(f_1 \land f_2) = \cost(f_1) \cdot \cost(f_2)$.
\end{lemma}

\begin{lemma}[Disjunction over Disjoint Domains]
  For separable cost function $\cost$ and Boolean functions $f_1$ and $f_2$ such that $\interpset{f_1} \cap \interpset{f_2} = \emptyset$:
    $\cost(f_1 \lor f_2) = \cost(f_1) + \cost(f_2)$.
\end{lemma}

@@

\begin{theorem}
\label{theorem:count}
Letting $n = |X|$, the following recursive formula holds for any node in a VPITE graph:
\begin{eqnarray}
\mcount(\tautology) & = & 2^{n} \label{eqn:count:true}\\
\mcount(\nil) & = & 0 \label{eqn:count:false}\\
\mcount(x) & = & 2^{n-1} \label{eqn:count:variable}\\
\mcount[\ite(v_i, v_t, v_e)] & = & \frac{\mcount(v_i) \cdot \mcount(v_t) + [2^n-\mcount(v_i)] \cdot \mcount(v_e)}{2^n} \label{eqn:count:ite}
\end{eqnarray}  
\end{theorem}
\noindent
{\em Proof:} The key result here is (\ref{eqn:count:ite}).  It follows
from Lemmas~\ref{lemma:negation}--\ref{lemma:sum}, along with the definition of the ITE operator.
